---
title: "Traffic Data and Map Plotting"
output: html_document
date: "2024-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if(!require(tidyverse)){
    install.packages("tidyverse")
    library(tidyverse)
}
if(!require(sf)){
    install.packages("sf")
    library(sf)
}
if(!require(censusapi)){
    install.packages("censusapi")
    library(censusapi)
}
```
Let's set up NY state shapefiles and plot them here in R. We'll plot them at several different levels of detail, depending on what the Census data gives.

First we load in the shapefile for Census block groups.
```{r}
nyc_cb_shapes <- st_read("../data/raw/nycb2020wi.shp")
```
We can view this as a geodataframe.

```{r}
View(nyc_cb_shapes)
```
## Making Plots

Plots in the `sf` package are straightforward.

```{r pressure, echo=FALSE}
plot(st_geometry(nyc_cb_shapes))
```
This can be annotated with colors. This code chooses the colors randomly.
```{r}
plot(st_geometry(nyc_cb_shapes), col = sf.colors(12, categorical = TRUE), border = 'grey', 
     axes = TRUE)
```
Notice that we can use `ggplot()` as well.
```{r}
ggplot()+geom_sf(data = nyc_cb_shapes)
```
We can also plot specific variables.
```{r}
plot(nyc_cb_shapes['BoroName'])
```
To make our life simpler going forward, let's only focus on Manhattan.
```{r}
manhattan_cb_shapes <- nyc_cb_shapes %>% filter(BoroName=="Manhattan")
```
# US Census API
Now let's bring in the US Census API for data from the American Community Survey. 

First, let's initialize the API key in our RStudio environment. Anyone can get a key but you can use mine. We should make people get their own.

```{r}
Sys.setenv(CENSUS_KEY="your_api_key_here")

Sys.getenv("CENSUS_KEY")
```
## List of Census Data Endpoints
We can see a full list of the data sources offered with the following commmand:

```{r}
apis <- listCensusApis()
```
Looking at this list, the endpoints which contain block- and tract-level data are the Planning Database datasets. These look like `pdb/~` where the second parameter could be tract, block, etc.

## State and County Codes

We also need state and county codes. You can find them here, in the drop-down menu titled "County and County Equivalent Entities".

https://www.census.gov/library/reference/code-lists/ansi.html

Since this file is working with NY, its codes are in this text file:

https://www2.census.gov/geo/docs/reference/codes2020/cou/st36_ny_cou2020.txt

Now we can make API calls. We will almost exclusively use the PDB data, so the `name` and `vintage` parameters will stay the same.

Eventually we'll want to select a whole list of variables. The full list of variables in the PDB is available, there almost 30,000 of them! You can see the all here. It will take a few seconds to pull down all the data.
```{r}
pdb_variables <- listCensusMetadata(name = "pdb/blockgroup", type = "variables", vintage=2020)
```

