---
title: "Easycensus Income Data and Interactive Maps"
output: html_document
date: "2024-01-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This notebook demonstrates how to pull demographic income data using the `easycensus` package, join it to the New York spatial bounds, and visualize the data interactively using `leaflet`.

## 1. Setup and Libraries

```{r load-libraries}
library(tidyverse)
library(sf)
library(xml2)
library(censusapi)
library(easycensus)
library(infer)
library(leaflet)

# Sys.setenv(CENSUS_KEY="your_api_key_here")
```

## 2. Prepare Base Spatial Data

First, we load the New York shapefiles and filter them down to Manhattan. Then, we calculate the number of citations per census tract based on our previously geocoded results.

```{r load-spatial}
# Read the NY Tracts shapefile
ny_ct_shapes <- st_read("../data/raw/ny_2010_t_bound.shp")

# Read combined geocoded results
df <- read.csv("../data/processed/Combined_Geocode_Results.csv")

# Filter and parse coordinates
geocoded_aggregate_df <- df %>% 
  filter(grepl(",", X_Y)) %>%  
  separate_wider_delim(X_Y, delim = ",", names = c("latitude", "longitude"))

# Convert to spatial object
df_spatial_coords_df <- st_as_sf(geocoded_aggregate_df, 
                                 coords = c("latitude", "longitude"), 
                                 crs = 4269, 
                                 agr = "constant")

# Filter base map to Manhattan
manhattan_ct_shapes <- ny_ct_shapes %>% filter(COUNTYFP == "061")

# Transform coordinates
df_spatial_coords_df <- st_transform(df_spatial_coords_df, crs = st_crs(manhattan_ct_shapes))

# Calculate citation counts
inter <- st_intersects(manhattan_ct_shapes, df_spatial_coords_df)
manhattan_ct_shapes$count <- lengths(inter)

# Plot simple heatmap of citations
ggplot(manhattan_ct_shapes) + 
  geom_sf(aes(fill = count)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Citation Count Heatmap")
```

## 3. Pulling Data with Easycensus

The `easycensus` package simplifies fetching ACS data. Let's pull median household income quintiles.

```{r easycensus-pull}
# Find variable keys
# cens_find_acs("income")

# Fetch data for B19081 (Income Quintiles) in New York County (Manhattan)
acs_income <- cens_get_acs("B19081", county = "New York", state = "NY", geography = "tract")

# Filter for the 3rd (middle) quintile
acs_income <- acs_income %>% filter(mnhiq_sub == "third quintile")

# Join income data to our shapefile
manhattan_income_shapes <- left_join(manhattan_ct_shapes, acs_income, by = join_by(GEOID == GEOID)) %>% 
  mutate(estimate = as.numeric(estimate))
```

### Income Distribution

Incomes are often right-skewed. Let's look at the distribution of our pulled estimates.

```{r plot-histogram}
manhattan_income_shapes %>%  
  ggplot(aes(x = estimate)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of 3rd Quintile Incomes", x = "Income Estimate", y = "Count")
```

### Static Overlay Map

We can plot a static map showing the income gradient overlaid with the citation coordinates (in red).

```{r static-overlay}
ggplot() +
  geom_sf(data = manhattan_income_shapes, aes(fill = estimate)) +
  geom_sf(data = df_spatial_coords_df, color = "red", size = 0.15, alpha = 0.5) +
  coord_sf(xlim = c(-73.93, -73.99), ylim = c(40.8, 40.84), expand = FALSE) +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal() +
  labs(title = "Traffic Citations and Annual Income, Upper Manhattan", fill = "Average Annual Income")
```

## 4. Interactive Maps with Leaflet

Leaflet allows us to create beautiful, interactive maps. Note that Leaflet requires the spatial data to be in the `WGS84` (EPSG:4326) coordinate reference system.

```{r prep-leaflet}
# Transform to EPSG:4326
manhattan_income_shapes <- st_transform(manhattan_income_shapes, 4326)
```

### Building the Base Interactive Map

```{r build-leaflet}
# Create the base map with income polygons
income_map <- leaflet(manhattan_income_shapes) %>% 
  addTiles() %>% 
  addPolygons(
    color = "#444444", 
    weight = 1, 
    smoothFactor = 0.5,
    opacity = 1.0, 
    fillOpacity = 0.5,
    fillColor = ~colorQuantile("YlOrRd", estimate)(estimate),
    highlightOptions = highlightOptions(
      color = "red", 
      weight = 2,
      bringToFront = TRUE
    )
  )

# View map
# income_map
```

### Subsampling Citations for Performance

Plotting all ~200,000 citations will crash the browser. We will subsample 100 citations to visualize them as markers on the map.

```{r leaflet-markers}
# Transform citations to WGS84
df_spatial_coords_df <- st_transform(df_spatial_coords_df, 4326)

# Sample 100 random citations
subsampled_citations <- rep_sample_n(df_spatial_coords_df, 100, reps = 1, replace = FALSE) %>% 
  mutate(
    lon = sf::st_coordinates(geometry)[, 1], 
    lat = sf::st_coordinates(geometry)[, 2]
  )

# Add markers to map
income_citations_map <- income_map %>% 
  addMarkers(
    data = subsampled_citations, 
    popup = ~ReturnAddress
  )

# Display final interactive map
income_citations_map
```
