---
title: "Traffic Data and Map Plotting (NYC)"
output: html_document
date: "2024-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(sf)
library(censusapi)

# Sys.setenv(CENSUS_KEY="your_api_key_here")
```

# Introduction

This notebook demonstrates how to load New York State shapefiles, filter them down to Manhattan, and plot Census demographic data over those geographic boundaries using the `sf` package.

## 1. Loading Shapefiles

First, we load the shapefile for Census tracts in New York City. We can view it as a geodataframe.

```{r load-shapefile}
# Load New York City Census Tracts shapefile
nyc_ct_shapes <- st_read("../data/raw/nyct2020wi.shp")

# View the structure of the spatial dataframe (optional)
# head(nyc_ct_shapes)
```

## 2. Making Spatial Plots

Plots in the `sf` package are straightforward. You can simply call `plot()` on the geometry.

```{r basic-plot}
# Basic geometry plot
plot(st_geometry(nyc_ct_shapes))
```

This can be annotated with colors. The following code chooses colors randomly for each tract.

```{r colored-plot}
plot(st_geometry(nyc_ct_shapes), col = sf.colors(12, categorical = TRUE), border = 'grey', axes = TRUE)
```

Notice that we can also use `ggplot()` and `geom_sf()`, which often produces cleaner graphics.

```{r ggplot-map}
ggplot() + 
  geom_sf(data = nyc_ct_shapes) +
  theme_minimal()
```

We can also plot specific variables contained within the shapefile attributes.

```{r variable-plot}
plot(nyc_ct_shapes['BoroName'])
```

## 3. Filtering by Borough

To make our analysis simpler and our maps faster to render, let's only focus on Manhattan.

```{r filter-manhattan}
manhattan_ct_shapes <- nyc_ct_shapes %>% filter(BoroName == "Manhattan")
```

# US Census API Integration

Now let's bring in the US Census API for data from the American Community Survey (ACS). 
First, ensure you have your API key initialized in your RStudio environment using `Sys.setenv(CENSUS_KEY="your_key")`.

## List of Census Data Endpoints

We can see a full list of the data sources offered with the following command:

```{r list-apis, eval=FALSE}
apis <- listCensusApis()
```

Looking at this list, the endpoints which contain block- and tract-level data are the **Planning Database** datasets. These look like `pdb/~`.

## Fetching Demographic Variables

For our purposes, we'll want extremely detailed data at the level of census tracts. We will request the `PUB_ASST_INC_ACS_14_18` variable (Public Assistance Income) for tracts in Manhattan (State 36, County 061).

```{r fetch-census-data}
# Pulling the variable for Public Assistance Income
pdb_pub_ast <- getCensus(
    name = "pdb/tract",
    vintage = 2020, 
    vars = c("PUB_ASST_INC_ACS_14_18", "Tract"),
    region = "tract:*",
    regionin = "state:36+county:061"
)

# Preview the data
head(pdb_pub_ast)
```

## Joining Data to Shapefiles

Now we join our retrieved Census demographic data to our Manhattan spatial boundaries using the Tract ID.

```{r join-data}
# Join the spatial dataframe with the census data
manhattan_pub_ast_shapes <- left_join(manhattan_ct_shapes, pdb_pub_ast, by = join_by(CT2020 == Tract))
```

Now we can plot the public assistance income data across the tracts in Manhattan!

```{r plot-demographics}
ggplot() + 
  geom_sf(data = manhattan_pub_ast_shapes, aes(fill = PUB_ASST_INC_ACS_14_18)) +
  scale_fill_viridis_c(na.value = "grey80") +
  theme_minimal() +
  labs(title = "Public Assistance Income in Manhattan", fill = "Assistance Level")
```

## Handling Missing Data

Sometimes census tracts won't have matching data (e.g. parks, airports, or unpopulated areas). We can find which tracts are missing using `setdiff`.

```{r check-missing}
unique_tracts_ct2020 <- unique(manhattan_ct_shapes$CT2020)
unique_tracts_tract <- unique(pdb_pub_ast$Tract)

# Find the difference - tracts in the shapefile but missing from Census data
missing_tracts <- setdiff(unique_tracts_ct2020, unique_tracts_tract)

# Display the missing tracts
print(missing_tracts)
```
