---
title: "Traffic Data Rhime Functions"
output: html_document
date: "2024-01-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This notebook demonstrates how to clean traffic parking data, geocode the addresses using the US Census API, and plot the citations on a map of Manhattan by census tract.

## 1. Environment Setup

First, we load the required libraries. If you don't have these installed, you can install them using `install.packages()`.

```{r load-libraries}
library(tidyverse)
library(sf)
library(xml2)
library(httr2)

# Load custom helper functions
source("../R/geocoding_helpers.R")
```

## 2. Load Spatial Boundaries

We load the 2010 New York census tract shapefiles to build our map.

```{r load-shapefiles}
# Read in the NY Tracts shapefile
ny_ct_shapes <- st_read("../data/raw/ny_2010_t_bound.shp")

# Quick plot to verify boundaries
ggplot() + 
  geom_sf(data = ny_ct_shapes) +
  theme_minimal() +
  labs(title = "New York 2010 Census Tracts")
```

## 3. Data Processing and Geocoding

Next, we load the parking citations data and prepare it for geocoding. 
*Note: The following code block is set to `eval=FALSE` so it doesn't run every time you knit the notebook. Running the full batch geocoding takes a long time.*

```{r batch-geocoding, eval=FALSE}
# Read the raw parking citations CSV
parking_csv <- "../data/raw/Parking_Fiscal_Year_2021.csv"
df <- read.csv(parking_csv)

# Create a full street address by combining House Number and Street Name
df$street_address <- paste(df[, 24], df[, 25], sep = " ")

# Use our helper function to batch geocode the addresses and save to data/processed
batch_geocode_addresses(df, chunk_size = 10000, output_dir = "../data/processed")
```

### 3.1 Combine Geocoded Results

After batch geocoding, we merge the chunk files into a single dataset.
*Note: This is also set to `eval=FALSE` because the combined file has already been generated.*

```{r combine-chunks, eval=FALSE}
# Read the first chunk
df_combined <- read.csv("../data/processed/Geocode_Result_Chunk_1.csv")
names(df_combined) <- c("UniqueID", "Address", "Matched", "Exactmatch", "ReturnAddress", "X_Y", "RandomID", "RValue")

# Loop through remaining chunks and row-bind
for(i in 2:19) {
  file_name <- paste0("../data/processed/Geocode_Result_Chunk_", i, ".csv")
  temp_df <- read.csv(file_name)
  names(temp_df) <- c("UniqueID", "Address", "Matched", "Exactmatch", "ReturnAddress", "X_Y", "RandomID", "RValue")
  df_combined <- rbind(df_combined, temp_df)
}

# Export combined dataframe
write.csv(df_combined, "../data/processed/Combined_Geocode_Results.csv", row.names = FALSE)
```

## 4. Spatial Joins

Now we load the combined geocoded results and convert them into spatial points so we can overlay them on the census tracts.

```{r process-coordinates}
# Read the combined geocoded results
df <- read.csv("../data/processed/Combined_Geocode_Results.csv")

# Filter rows with coordinates and split into latitude and longitude
geocoded_aggregate_df <- df %>% 
  filter(grepl(",", X_Y)) %>%  
  separate_wider_delim(X_Y, delim = ",", names = c("latitude", "longitude"))

# Convert to spatial object
df_spatial_coords_df <- st_as_sf(geocoded_aggregate_df, 
                                 coords = c("latitude", "longitude"), 
                                 crs = 4269, 
                                 agr = "constant")
```

### 4.1 Filter to Manhattan

We'll focus our analysis on Manhattan (County FIPS code 061).

```{r filter-manhattan}
# Filter shapefile to Manhattan only
manhattan_ct_shapes <- ny_ct_shapes %>% filter(COUNTYFP == "061")

# Transform coordinates to match the shapefile CRS
df_spatial_coords_df <- st_transform(df_spatial_coords_df, crs = st_crs(manhattan_ct_shapes))
```

### 4.2 Calculate Citations per Tract

We intersect the citation coordinates with the census tracts to count how many citations occurred in each tract.

```{r calculate-counts}
# Count intersections
inter <- st_intersects(manhattan_ct_shapes, df_spatial_coords_df)

# Store the counts
manhattan_ct_shapes$count <- lengths(inter)
```

## 5. Heatmap Visualization

Finally, we plot a heatmap showing the density of traffic citations across Manhattan.

```{r plot-heatmap}
ggplot(manhattan_ct_shapes) + 
  geom_sf(aes(fill = count), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", trans = "log1p", breaks = c(0, 10, 100, 1000)) +
  theme_minimal() +
  labs(
    title = "Traffic Citations in Manhattan",
    subtitle = "By Census Tract",
    fill = "Citations"
  )
```

## 6. Optional: Demographic Relationships

You can also use the Census API to download demographic data (e.g., age demographics) and see if they correlate with citation frequency.

```{r load-census-api}
# Ensure your census key is set
# Sys.setenv(CENSUS_KEY="your_key_here")

library(censusapi)

# Fetch Population 18-24 by tract
pdb_pub_ast <- getCensus(
    name = "pdb/tract",
    vintage = 2020, 
    vars = c("Pop_18_24_CEN_2010", "Tract"),
    region = "tract:*",
    regionin = "state:36+county:061"
)

# Join demographic data to shapefile
manhattan_pub_ast_shapes <- left_join(manhattan_ct_shapes, pdb_pub_ast, by = join_by(TRACTCE == Tract))

# Plot the demographic variable
ggplot() + 
  geom_sf(data = manhattan_pub_ast_shapes, aes(fill = Pop_18_24_CEN_2010)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Population Aged 18-24 in Manhattan")

# Check relationship
lm_model <- lm(count ~ Pop_18_24_CEN_2010, data = manhattan_pub_ast_shapes)
summary(lm_model)
```
